---
title: "Final Project"
author: "Group 20"
date: "12/06/2023"
output:
  pdf_document: default
  html_document: default
---

```{r, message=FALSE}
# install.packages('NHANES')
library(NHANES)
library(dplyr)
library(ggplot2)
library(car)
```

# 2. Methods
**remove duplicates**  
```{r, message=FALSE}
data(NHANES)
NHANES <- NHANES[!duplicated(NHANES$ID), ]
dim(NHANES)
```

**data processing**  
```{r, message=FALSE}
# heavy drink
NHANES$AlcoholTotal <- NHANES$AlcoholDay * NHANES$AlcoholYear / 365
NHANES$heavydrink <- 'No'
NHANES$heavydrink[(NHANES$Gender=='male')&(NHANES$AlcoholTotal>=2)] <- 'Yes'
NHANES$heavydrink[(NHANES$Gender=='female')&(NHANES$AlcoholTotal>=1)] <- 'Yes'
# drinking days per year / 100
NHANES$DrinkingDaysPerYear <- NHANES$AlcoholYear / 100
```

```{r, message=FALSE}
# Age: centered
NHANES$AgeC <- NHANES$Age - mean(NHANES$Age)
# Race: set White as reference
NHANES$Race1 <- relevel(NHANES$Race1, ref='White')
# Poverty: not poverty as reference group
NHANES$Poverty1 <- 5 - NHANES$Poverty
# Depressed: regroup
NHANES$Depressed1 <- ifelse(NHANES$Depressed=='None', 'No',
  ifelse(NHANES$Depressed %in% c('Several', 'Most'), 'Yes', NA))
```

**remove non-drinkers & NA**  
```{r, message=FALSE}
NHANES1 <- NHANES[NHANES$AlcoholYear > 0, ]
NHANES1 <- na.omit(NHANES1[, c('DirectChol', 'AgeC', 'Race1', 'Gender', 'Poverty1', 'Education', 'BMI_WHO', 'Diabetes', 'Smoke100', 'Depressed1', 'PhysActive',
  'AlcoholYear', 'DrinkingDaysPerYear', 'heavydrink')])
dim(NHANES1)
```


summary(NHANES1$Race1)

**Initial Model**  
```{r, message=FALSE}
model.initial <- lm(DirectChol ~ AgeC + Gender + Race1 + 
  Poverty1 + Education +
  BMI_WHO + Diabetes + Smoke100 + Depressed1 + PhysActive +
  AlcoholYear + heavydrink +
  AlcoholYear*Gender + heavydrink*Gender +
  AlcoholYear*AgeC + heavydrink*AgeC +
  AlcoholYear*Poverty1 + heavydrink*Poverty1 +
  AlcoholYear*Race1 + heavydrink*Race1, data = NHANES1)
```

```{r, message=FALSE}
par(mfrow = c(2, 2))
plot(model.initial)
# histogram before log
ggplot(data=NHANES1) + geom_histogram(mapping = aes(x = DirectChol))
# histogram after log
ggplot(data=NHANES1) + geom_histogram(mapping = aes(x = log(DirectChol)))
```
1. QQ-plot => residual not normally distributed
2. residual plots => not constant variance
3. histogram of Y => right skewed
=> log tranformation to Y

# Model
```{r, message=FALSE}
model <- lm(log(DirectChol) ~ AgeC + Gender + Race1 + 
  Poverty1 + Education +
  BMI_WHO + Diabetes + Smoke100 + Depressed1 + PhysActive +
  DrinkingDaysPerYear + heavydrink + 
  DrinkingDaysPerYear*AgeC + heavydrink*AgeC +
  DrinkingDaysPerYear*Poverty1 + heavydrink*Poverty1, data = NHANES1)
summary(model)
```


# Model Diagnostics
**LINE**  
```{r}
# normality
car::qqPlot(model)
# constant variance
par(mfrow = c(2, 2))
plot(model)
```
1. no clear pattern of the distribution of residuals over the predicted range => constant variance  
2. QQ-plot => normality  

```{r}
# linearity
car::residualPlots(model, type="response")
```
p-value > 0.05, not reject H0, linearity

```{r}
# independence
durbinWatsonTest(model)
```
p-value > 0.05, not reject H0, rho=0, independent

**multicollinearity**  
```{r}
# multicollinearity
car::vif(model, type='predictor')
```
From the GVIF^(1/(2Df)) values in the output, none of the values exceed sqrt(5), which is equivalent to VIF < 5, we do not observe any potential collinearity issues.

**influence diagnostics**  

```{r}
library(olsrr)
influence = data.frame(Residual = resid(model), Rstudent = rstudent(model),
HatDiagH = hat(model.matrix(model)),
CovRatio = covratio(model), DFFITS = dffits(model),
COOKsDistance = cooks.distance(model))

ols_plot_dffits(model)
influence[order(abs(influence$DFFITS),decreasing = T),] %>% head()
```

```{r}
ols_plot_cooksd_bar(model)
influence[order(influence$COOKsDistance,decreasing = T),] %>% head()
```

```{r}
ols_plot_resid_lev(model)
# high leverage
influence[order(influence$HatDiagH,decreasing = T),] %>% head()
# high studentized residual
influence[order(influence$Rstudent,decreasing = T),] %>% head()
```

**refit the model after removing outliers**  
```{r}
obs <- rownames(influence[(abs(influence$HatDiagH)>0.019) & (abs(influence$Rstudent)>2),])
obs <- c(obs, c('888', '2109', '1008', '885'))
NHANES1.rm <- NHANES1[-as.numeric(obs), ]
model.rm <- lm(log(DirectChol)~AgeC + Gender + Race1 + 
  Poverty1 + Education +
  BMI_WHO + Diabetes + Smoke100 + Depressed1 + PhysActive +
  DrinkingDaysPerYear + heavydrink + 
  DrinkingDaysPerYear*AgeC + heavydrink*AgeC +
  DrinkingDaysPerYear*Poverty1 + heavydrink*Poverty1, data = NHANES1.rm)
summary(model.rm)
```
no obvious difference




```{r}
library(gridExtra)
library(broom)

plots <- list()

plot <- ggplot(data=NHANES1) + 
  geom_histogram(mapping = aes(x = DirectChol), bins = 21) + 
  ggtitle(paste("Distribution of Response Variable"))
plots[[1]] <- plot

plot <- ggplot(data=NHANES1) + 
  geom_histogram(mapping = aes(x = log(DirectChol)), bins = 21) + 
  ggtitle(paste("Distribution of Log Transformed Response Variable"))
plots[[2]] <- plot

plot <- ggplot(augment(model.initial), aes(.fitted, sqrt(abs(.resid)))) +
  geom_point() +
  geom_smooth(se = FALSE, method = "loess", color = "blue") + 
  ggtitle("Scale-Location Plot for Non-Logged Response Variable") +
  theme_minimal()
plots[[3]] <- plot

plot <- ggplot(augment(model), aes(.fitted, sqrt(abs(.resid)))) +
  geom_point() +
  geom_smooth(se = FALSE, method = "loess", color = "blue") + 
  ggtitle("Scale-Location Plot for Logged Response Variable") +
  theme_minimal()
plots[[4]] <- plot

plot <- ggplot(augment(model.initial), aes(sample = .resid)) +
    geom_qq() +
    stat_qq_line(linetype = "dotdash", color = "red") + 
    ggtitle("QQ Plot for Non-Logged Residuals") +
    theme_minimal()
plots[[5]] <- plot

plot <- ggplot(augment(model), aes(sample = .resid)) +
    geom_qq() +
    stat_qq_line(linetype = "dotdash", color = "red") + 
    ggtitle("QQ Plot for Logged Residuals") +
    theme_minimal()
plots[[6]] <- plot

grid.arrange(grobs = plots, ncol = 2)

```

```{r}
plots <- list()

# linearity
# car::residualPlots(model, type = "response", subset = ~ AgeC + Poverty1 + DrinkingDaysPerYear)

data_subset <- data.frame(AgeCentered = NHANES1$AgeC, Poverty = NHANES1$Poverty1, DrinkingDaysPerYear = NHANES1$AlcoholYear, residuals = residuals(model))

plot <- ggplot(data_subset, aes(x = AgeCentered, y = residuals)) +
  geom_point(shape = 1) +
  geom_smooth() +
  labs(title = "Residuals for AgeCenter - Linearity") + 
  theme(plot.title = element_text(size = 12))
plots[[1]] <- plot

plot <- ggplot(data_subset, aes(x = Poverty, y = residuals)) +
  geom_point(shape = 1) +
  geom_smooth() +
  labs(title = "Residuals for Poverty - Linearity") +
  theme(plot.title = element_text(size = 12))
plots[[2]] <- plot

plot <- ggplot(data_subset, aes(x = DrinkingDaysPerYear, y = residuals)) +
  geom_point(shape = 1) +
  geom_smooth() +
  labs(title = "Residuals for DrinkingDaysPerYear - Linearity") +
  theme(plot.title = element_text(size = 12))
plots[[3]] <- plot


# Equal Variance
plot <- ggplot(augment(model), aes(.fitted, sqrt(abs(.resid)))) +
  geom_point() +
  geom_smooth(se = FALSE, method = "loess", color = "blue") + 
  ggtitle("Scale-Location Plot - Equal Variance") +
  theme(plot.title = element_text(size = 12))
plots[[4]] <- plot

# Normality
plot <- ggplot(augment(model), aes(sample = .resid)) +
    geom_qq() +
    stat_qq_line(linetype = "dotdash", color = "blue") + 
    ggtitle("QQ Plot - Normality") +
    theme(plot.title = element_text(size = 12))
plots[[5]] <- plot


plot <- ggplot(augment(model), aes(.fitted, .resid)) +
  geom_point() +
  ggtitle("Residuals vs Fitted") +
  geom_abline(slope = 0, intercept = 0, linetype = "dashed", color = "blue") +
  theme(plot.title = element_text(size = 12))
plots[[6]] <- plot

grid.arrange(grobs = plots, ncol = 3)

```

